---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.cap = "",
  out.width = "100%",
  fig.width = 8,
  fig.height = 4.2,
  dpi = 150
)
```

# dfmtools

<!-- badges: start -->
[![R-CMD-check](https://github.com/franzmohr/dfmtools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/franzmohr/dfmtools/actions/workflows/R-CMD-check.yaml)
[![License: GPL v2](https://img.shields.io/badge/License-GPL%20v2-blue.svg)](https://www.gnu.org/licenses/old-licenses/gpl-2.0.en.html)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Bayesian inference of dynamic factor models. `dfmtools` builds the model object,
attaches priors and starting values, and hands the whole thing to a Gibbs
sampler written in C++, following the treatments in Chan, Koop, Poirier and
Tobias (2019) and Lütkepohl (2006).

The model is a measurement equation

$$x_t = \lambda f_t + u_t, \qquad u_t \sim N(0, U),$$

for an $M \times 1$ vector of observed variables $x_t$ and $N \times 1$
unobserved factors $f_t$, together with a transition equation

$$f_t = \sum_{i=1}^{p} A_i f_{t-i} + v_t, \qquad v_t \sim N(0, V).$$

Five blocks are drawn in turn: the path of the factors, the loadings, the two
error precisions and the transition coefficients. `U` and `V` are diagonal and
estimated with independent gamma priors on their precisions.

## Installation

`dfmtools` needs `bvartools (>= 1.0.0)`, which is not on CRAN yet, so both come
from GitHub:

```r
# install.packages("remotes")
remotes::install_github("franzmohr/bvartools")
remotes::install_github("franzmohr/dfmtools")
```

A C++17 compiler and GNU make are required to build from source — Rtools on
Windows, the Xcode command line tools on macOS.

## Usage

A model is specified, given priors and starting values, and then drawn from.
Each step returns the same object with one more element on it.

```{r setup, message = FALSE}
library(dfmtools)

data("bem_dfmdata")

dim(bem_dfmdata)
```

`bem_dfmdata` is 225 quarters of 196 US macroeconomic series, taken from the data
sets accompanying Chan et al. (2019). Two factors and a transition of order two:

```{r specify}
model <- create_dfmodel(x = bem_dfmdata, p = 2, n = 2,
                        iterations = 5000, burnin = 1000)

model <- add_priors(model,
                    lambda = list(vinv = 0.01),
                    u = list(shape = 5, rate = 4),
                    a = list(vinv = 0.01),
                    v = list(shape = 5, rate = 4))

model <- add_initial_values(model)
```

`create_dfmodel` normalises every column of `x` by default, so the loadings are
comparable across series. Priors on `lambda` and `a` are specified as
*precisions*; `u` and `v` take the shape and rate of the gamma prior on the error
precisions.

```{r draw}
set.seed(6023)

model <- add_posterior_coefficients(model)

vapply(model$posterior, function(x) paste(dim(x$coeffs), collapse = " x "), "")
```

Draws run along the rows, so a posterior mean is a column mean. Each block is a
`coda::mcmc` matrix, and the whole of a `T`-period factor path is part of every
draw: the factors are unobserved, so neither the forecast nor the log-likelihood
can be recomputed without it.

Only the product $\lambda f_t$ is identified, so the leading $N \times N$ block
of the loading matrix is fixed unit lower triangular rather than drawn. It is
reported all the same, so that a draw is reshaped rather than unpacked:

```{r loadings}
lambda <- matrix(colMeans(model$posterior$lambda$coeffs),
                 nrow = model$model$m, ncol = model$model$n,
                 dimnames = list(colnames(model$data$x), c("f1", "f2")))

round(head(lambda, 6), 3)
```

The factor path is stored period by period, all `N` factors of a period
together, which makes a single factor a strided slice:

```{r factors}
tt <- nrow(model$data$x)
n <- model$model$n

factor_draws <- function(i) model$posterior$factors$coeffs[, seq(i, tt * n, by = n)]

f1 <- apply(factor_draws(1), 2, quantile, probs = c(0.1, 0.5, 0.9))
dim(f1)
```

```{r factor-path, echo = FALSE, fig.alt = "Posterior median and 80% credible interval of the two factors, 1959Q3 to 2015Q3."}
band <- function(q, ...) {
  polygon(c(time(model$data$x), rev(time(model$data$x))), c(q[1, ], rev(q[3, ])),
          border = NA, ...)
}

f2 <- apply(factor_draws(2), 2, quantile, probs = c(0.1, 0.5, 0.9))
tm <- as.numeric(time(model$data$x))

par(mar = c(3, 3, 2, 1), mgp = c(2, 0.6, 0), las = 1)
plot(range(tm), range(f1, f2), type = "n", xlab = "", ylab = "",
     main = "Posterior of the two factors (median, 80% interval)")
abline(h = 0, col = "grey80")
band(f1, col = grDevices::adjustcolor("#1f4e79", alpha.f = 0.2))
band(f2, col = grDevices::adjustcolor("#a33e00", alpha.f = 0.2))
lines(tm, f1[2, ], col = "#1f4e79", lwd = 1.8)
lines(tm, f2[2, ], col = "#a33e00", lwd = 1.8)
legend("topleft", legend = c("Factor 1", "Factor 2"), bty = "n",
       col = c("#1f4e79", "#a33e00"), lwd = 1.8)
```

### Forecasts

A dynamic factor model needs no out-of-sample regressors, so only the horizon has
to be supplied. Each draw's path is the transition run forward from the last `p`
drawn factors with an innovation at every step, and the observable variables are
read off the loadings — both error terms drawn, so the result is from the
posterior predictive distribution rather than a conditional mean.

```{r forecast}
model <- add_posterior_forecasts(model, n_ahead = 8)

dim(model$posterior$forecast)
```

The horizons are stacked within a row in the variable order of the sample, so
reshaping a draw to $M \times h$ puts the variables in the rows:

```{r forecast-gdp}
gdp <- which(colnames(model$data$x) == "GDPC96")

fc <- t(apply(model$posterior$forecast[, seq(gdp, 8 * model$model$m,
                                            by = model$model$m)],
              2, quantile, probs = c(0.1, 0.5, 0.9)))
rownames(fc) <- paste0("h = ", 1:8)

round(fc, 3)
```

Those are in the units of the normalised data. Multiplying by the sample standard
deviation of the series and adding its mean back returns them to the original
scale.

### Log-likelihood

`add_posterior_loglik` returns the pointwise log-likelihood laid out for WAIC and
PSIS-LOO: draws along the rows, observations along the columns. It is the
measurement density at the drawn factors, so it is the log-likelihood
*conditional* on the factor path rather than the marginal one — which matters for
what an information criterion computed from it means.

```{r loglik}
model <- add_posterior_loglik(model)

dim(model$posterior$loglik)

# WAIC, computed on a log scale so that 196 variables do not underflow.
ll <- model$posterior$loglik
logmeanexp <- function(z) {
  mx <- max(z)
  mx + log(mean(exp(z - mx)))
}

lppd <- sum(apply(ll, 2, logmeanexp))
p_waic <- sum(apply(ll, 2, stats::var))

round(-2 * (lppd - p_waic), 1)
```

### Several specifications at once

Integer vectors for `p` or `n` produce one model per combination, and the rest of
the chain maps over them:

```{r modellist}
models <- create_dfmodel(x = bem_dfmdata, p = 1:2, n = 1:2,
                         iterations = 200, burnin = 100)

length(models)

vapply(models, function(z) sprintf("p = %d, n = %d", z$model$p, z$model$n), "")
```

## Where the sampler lives

Posterior simulation is not implemented in R. `src/core/` and
`inst/include/bayests/` are vendored copies of part of the core layer of
[BayesTS](https://github.com/franzmohr/BayesTS), and this package is a
translation layer over it — the same arrangement `bvartools` uses for its VAR and
VEC models. `src/core/VENDORED.md` records which files are copied, how that set
is computed from the `#include` graph, and the one modification applied on the
way in.

The practical consequence for a user is that `set.seed()` reaches the sampler:
Armadillo's RNG is R's own.

## Getting help

* Bugs and feature requests: [issues](https://github.com/franzmohr/dfmtools/issues)
* Contributing: [.github/CONTRIBUTING.md](.github/CONTRIBUTING.md)
* Numerical issues in the sampler itself belong in
  [BayesTS](https://github.com/franzmohr/BayesTS/issues)

## References

Chan, J., Koop, G., Poirier, D. J., & Tobias, J. L. (2019). *Bayesian
Econometric Methods* (2nd ed.). Cambridge: Cambridge University Press.

Lütkepohl, H. (2006). *New Introduction to Multiple Time Series Analysis* (2nd
ed.). Berlin: Springer.

McCracken, M. W., & Ng, S. (2016). FRED-MD: A monthly database for macroeconomic
research. *Journal of Business & Economic Statistics*, 34(4), 574-589.

## License

GPL (>= 2). The vendored BayesTS files carry BSD-3-Clause headers, which is
compatible; `inst/COPYRIGHTS` lists them file by file and also notes that
`bem_dfmdata` is third-party data whose terms are not this package's to grant.
