---
title: "Identifying a monetary policy shock"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Identifying a monetary policy shock}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.4,
  dpi = 120
)
```

A factor augmented VAR exists to answer a question a small VAR answers badly.
Estimate the effect of monetary policy in a three- or five-variable VAR and the
price level rises after a contractionary shock -- the price puzzle -- which is
not a result but a symptom: the central bank sees more than the three variables
do, so what the VAR reads as a policy shock is partly the bank reacting to
information the VAR never saw. Bernanke, Boivin and Eliasz (2005) put the
information back by summarising a large panel into a few factors and putting
those in the VAR beside the policy rate.

This vignette follows their design on the data that ships with the package. It
is their *design*, not their numbers: they use 120 monthly series through
2001:8, and `bem_dfmdata` is quarterly.

```{r setup}
library(dfmtools)
data("bem_dfmdata")
```

## The model

The panel is a measurement equation on the whole state,

$$x_t = \lambda_f f_t + \lambda_y y_t + e_t, \qquad e_t \sim N(0, R),$$

with $R$ diagonal, and the state is a VAR in the factors and the observed
variables together,

$$s_t = \sum_{i=1}^{p} \Phi_i s_{t-i} + v_t, \qquad s_t = (f_t', y_t')', \qquad
v_t \sim N(0, Q).$$

The policy rate is $y_t$. It is part of the state rather than a regressor: it
appears on the left of the transition as well as the right, and $Q$ is
unrestricted so that its cross block -- the correlation between the factor
innovations and the shock to the policy rate -- is estimated rather than
assumed. That block is the whole reason to fit the model, and it is also why the
model cannot identify anything on its own. An unrestricted $Q$ has no preferred
rotation. The identification has to come from outside.

## Ordering the panel

Two restrictions do the work, and both are decisions about the order of the
columns of `x`.

The first is the normalisation `create_favarmodel()` imposes: the leading
$n \times n$ block of $\lambda_f$ is the identity and the observed columns of
those rows are zero. Factor $i$ is therefore panel series $i$ plus idiosyncratic
noise, exactly. That fixes the rotation, but it also means the first $n$ columns
of the panel are a substantive choice and not a formality.

The second is Bernanke, Boivin and Eliasz's own. Divide the panel into
*slow-moving* series -- output, employment, prices, which do not react to a
policy surprise within the period -- and *fast-moving* ones -- interest rates,
spreads, equity, the exchange rate, which reprice on the day. Slow-moving
series are restricted to load zero on the policy rate contemporaneously.
Combine that with the normalisation and the factors inherit the property: if
the series that define them cannot react within the period, neither can they.

So the identifying columns must be slow-moving ones, and it is worth picking
series that are broad rather than headline, since whatever goes first *becomes*
a factor.

```{r panel}
identifiers <- c("INDPRO", "PAYEMS", "CPIAUCSL")

slow <- c(identifiers,
          "GDPC96", "PCECC96", "IPFINAL", "USPRIV", "UNRATE",
          "AWHMAN", "PCECTPI", "GDPCTPI", "HOUST")

fast <- c("TB3MS", "GS1", "GS10", "GS10TB3Mx", "M1REALx", "M2REALx",
          "BUSLOANSx", "EXUSUKx", "S.P.500", "S.P.div.yield",
          "TB3SMFFM", "AAAFFM")

panel <- bem_dfmdata[, c(slow, fast)]
ffr <- bem_dfmdata[, "FEDFUNDS", drop = FALSE]
```

`bem_dfmdata` arrives with the FRED-QD transformations already applied, so the
real quantities and the price indices are growth rates. `FEDFUNDS` is not
transformed and stays in percentage points, which is what makes a response to it
readable later.

```{r data}
c(quarters = nrow(panel), series = ncol(panel))
round(summary(as.numeric(ffr)), 2)
```

## Estimation

Three factors and four lags, the quarterly counterpart of their three factors
and thirteen monthly lags.

```{r model}
model <- create_favarmodel(x = panel, y = ffr, p = 4, n = 3,
                           iterations = 4000, burnin = 1000)
```

`normalize_x = TRUE` is the default and divides each panel column by its own
standard deviation. Without it a factor extracted from series of different
scales is a factor of the largest of them. The consequence to remember is that a
panel response comes back in standard deviations of that series rather than in
its units.

Argument `slow` of `add_priors()` carries the second restriction. The
identifying three are listed with the rest, which is allowed and does nothing --
the normalisation has already zeroed their observed columns -- but says plainly
that they belong to the slow group, which is the condition the identification
rests on.

```{r priors}
model <- add_priors(model, slow = slow)
model <- add_initial_values(model)
```

The restriction is a prior held at a precision of `1e12`, not a hard zero. That
is close enough to be exact in practice, and it is worth checking rather than
assuming.

```{r estimate}
set.seed(20050201)
model <- add_posterior_coefficients(model)

lambda <- matrix(colMeans(model$posterior$lambda$coeffs),
                 ncol(panel), model$model$n + model$model$n_obs)

range(lambda[seq_along(slow), 4])          # slow block, loading on FEDFUNDS
range(lambda[-seq_along(slow), 4])         # fast block
```

The slow block is at zero to eight decimal places while the fast block is not,
which is the restriction doing what it was asked to.

## The shock

`irf()` identifies the shock with the Cholesky factor of $Q$ in the order given
by `order`, which defaults to the state's own: the factors, then the observed
block. The policy rate is therefore ordered last, so it may respond to the
factors within the period and they may respond to it only with a lag -- the
recursive assumption of Bernanke, Boivin and Eliasz, which the slow-moving
restriction is what makes credible.

```{r policy}
policy <- irf(model, impulse = "FEDFUNDS", response = "FEDFUNDS", n_ahead = 20)
plot(policy, main = "Federal funds rate")
```

One standard deviation of the orthogonalised innovation is a little over thirty
basis points on impact. Most of it is gone within two quarters, leaving a small
and persistent remainder -- and the band covers zero from the first quarter on,
which is worth registering before reading anything downstream of it.

## Reading the responses

The panel is in growth rates, so a response to be read as a level response has
to be accumulated. That is `cumulative = TRUE`, and forgetting it is the easiest
way to mistake a hump-shaped decline in the level of output for noise in its
growth rate.

```{r responses, fig.height = 6}
opar <- par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

for (v in c("GDPC96", "INDPRO", "PCECC96", "CPIAUCSL")) {
  plot(irf(model, impulse = "FEDFUNDS", response = v, n_ahead = 20,
           cumulative = TRUE),
       main = v, ylab = "Std. deviations")
}

par(opar)
```

The medians have the shape Bernanke, Boivin and Eliasz report. Output,
industrial production and consumption fall, trough about two years out and
recover, and the price level falls monotonically rather than rising, so there is
no price puzzle of the kind a small VAR on the same sample produces.

The bands are another matter, and they are the result as much as the medians
are: they cover zero at every horizon in all four panels. Twenty-four quarterly
series is a thin panel beside their 120 monthly ones, the sample runs through a
decade over which the funds rate could not move, and four lags of a four-wide
state is a small model to ask this of. The exercise shows that the machinery
reproduces their design and their qualitative pattern. It does not show that the
pattern is distinguishable from zero on this data.

Unemployment is a level already and is not accumulated.

```{r unemployment}
plot(irf(model, impulse = "FEDFUNDS", response = "UNRATE", n_ahead = 20),
     main = "Unemployment rate")
```

Note which of these are read through the loadings. `GDPC96`, `PCECC96` and
`UNRATE` are ordinary panel series that never enter the transition; their
responses are $\lambda_{j \cdot} \Psi_h P e_i$, the state's response carried
through their row of the loadings. That is what the factor structure buys -- a
response for every series in the panel, from a VAR of width four. `INDPRO` and
`CPIAUCSL` are identifying series, so what is plotted for them is the response
of factors one and three.

## What this is not

Two things separate the above from a replication of Bernanke, Boivin and Eliasz.

The data is the first. Theirs is monthly, 120 series, 1959:1 to 2001:8, and
stops before the zero lower bound; this is quarterly FRED-QD through 2015Q3,
which includes a long stretch over which the funds rate could not move. Their
panel is in their replication files and FRED-MD is the monthly successor, and
either drops into the code above unchanged.

The estimator is the second. They report two estimators, a two-step principal
components procedure and the one-step likelihood-based one. This package
implements the one-step: the factors, the loadings, the transition and both
error blocks are drawn jointly, so the uncertainty about the factors is carried
into the response bands rather than bootstrapped back in afterwards. The
two-step estimator is not available here.

## References

Bernanke, B. S., Boivin, J., & Eliasz, P. (2005). Measuring the effects of
monetary policy: A factor-augmented vector autoregressive (FAVAR) approach.
*The Quarterly Journal of Economics, 120*(1), 387--422.

McCracken, M. W., & Ng, S. (2016). FRED-MD: A monthly database for
macroeconomic research. *Journal of Business & Economic Statistics, 34*(4),
574--589.
